---
title: 'Inference for numerical data'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(oilabs)
```

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the `tidyverse` suite of 
packages, and perform statistical inference using `infer`. The data can be found
in the companion package for OpenIntro labs, `oilabs`.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(oilabs)
library(infer)
```

### Creating a reproducible lab report

To create your new lab report, start by opening a new R Markdown document... From Template... then select Lab Report from the `oilabs` package. 

### The data

Every two years, the Centers for Disease Control and Prevention conduct the Youth
Risk Behavior Surveillance System (YRBSS) survey, where it takes data from high 
schoolers (9th through 12th grade), to analyze health patterns. You will work 
with a selected group of variables from a random sample of observations during 
one of the years the YRBSS was conducted.


Load the `yrbss` data set into your workspace.

```{r load-data}
data(yrbss)
```

There are observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-nc}
?yrbss
```


1.  What are the cases in this data set? How many cases are there in our sample?

Remember that you can answer this question by viewing the data in the data viewer or
by using the following command:

```{r str}
glimpse(yrbss)
```

## Exploratory data analysis

You will first start with analyzing the weight of the participants in kilograms:
`weight`.

Using visualization and summary statistics, describe the distribution of weights.
The `summary` function can be useful.

```{r summary}
summary(yrbss$weight)
```

1.  How many observations are we missing weights from?

Next, consider the possible relationship between a high schooler's weight and their
physical activity. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

First, let's create a new variable `physical_3plus`, which will be coded as either
"yes" if they are physically active for at least 3 days a week, and "no" if not.

```{r create new var}
yrbss <- yrbss %>% 
  mutate(physical_3plus = ifelse(yrbss$physically_active_7d > 2, "yes", "no"))
```


2.  Make a side-by-side boxplot of `physical_3plus` and `weight`. Is there a 
relationship between these two variables? What did you expect and why?

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following to 
first group the data by the `physical_3plus` variable, and then calculate the mean
`weight` in these groups using the `mean` function while ignoring missing values
by setting the `na.rm` argument to `TRUE`.

```{r by-means}
yrbss %>%
  group_by(physical_3plus) %>%
  summarise(mean_weight = mean(weight, na.rm = TRUE))
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test.

## Inference

3.  Are all conditions necessary for inference satisfied? Comment on each. You can 
compute the group sizes with the `summarize` command above by defining a new variable
with the definition `n()`.

4.  Write the hypotheses for testing if the average weights are different for those
who exercise at least times a week and those who don't.

Next, we will introduce a new function, `hypothesize`, that falls into the `infer` 
workflow. You will use this method for conducting hypothesis tests. 

But first, we need to initialize the test, which we will save as `d_hat`.

```{r inf-weight-habit-ht-initial, tidy=FALSE}
d_hat <- yrbss %>%
  specify(weight ~ physical_3plus) %>%
  calculate(stat = "diff in means", order = c("yes", "no"))
```

Notice how you can use the functions `specify` and `calculate` again like you did
for calculating confidence intervals. Here, though, the statistic you are searching
for is the difference in means, with the order being `yes - no != 0`.

After you have initialized the test, you need to simulate the test on the null
distribution, which we will save as `null`.

```{r inf-weight-habit-ht-null, tidy=FALSE}
null <- yrbss %>%
  specify(weight ~ physical_3plus) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("yes", "no"))
```

Here, `hypothesize` is used to set the null hypothesis as a test for independence.
Also, note that the `type` argument within `generate` is set to `permute`, which 
is the argument when generating a null distribution for a hypothesis test.

Now that the test is initialized and the null distribution formed, you can calculate
the p-value for your hypothesis test using the function `get_p_value`.

```{r inf-weight-habit-ht-pvalue}
null %>%
  get_p_value(obs_stat = d_hat, direction = "two_sided")
```

This the standard workflow for performing hypothesis tests.

5.  Construct and record a confidence interval for the difference between the 
weights of those who exercise at least three times a week and those who don't, and
interpret this interval in context of the data.

* * *

## More Practice

6.  Calculate a 95% confidence interval for the average height in meters (`height`)
and interpret it in context.

7.  Calculate a new confidence interval for the same parameter at the 90% 
confidence level. Comment on the width of this interval versus 
the one obtained in the previous exercise.

8.  Conduct a hypothesis test evaluating whether the average height is different
for those who exercise at least three times a week and those who don't.

9.  Now, a non-inference task: Determine the amount of different options there 
are in the dataset for the `hours_tv_per_school_day` there are.

10. Come up with a research question evaluating the relationship between height 
or weight and sleep. Formulate the question in a way that it can be answered using
a hypothesis test and/or a confidence interval. Report the statistical results, 
and also provide an explanation in plain language. Be sure to check all 
assumptions, state your $\alpha$ level, and conclude in context.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab 
written by the faculty and TAs of UCLA Statistics.
</div>