---
title: "Inference for categorical data"
runtime: shiny
output:
  html_document:
    css: www/lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, message = FALSE, warning = FALSE)
```

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of 
packages, and perform statistical inference using **infer**. The data can be found
in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE, eval=TRUE}
library(tidyverse)
library(openintro)
library(infer)
```

### Creating a reproducible lab report

To create your new lab report, start by opening a new R Markdown document... From Template... then select Lab Report from the **openintro** package. 

### The data

You will be analyzing the same dataset as in the previous lab, where you delved 
into a sample from the Youth Risk Behavior Surveillance System (YRBSS) survey,
which uses data from high schoolers to help discover health patterns.

Load the `yrbss` data set into your workspace.

```{r load-data}
data(yrbss)
```

1.   What are the counts within each category for the amount of days these students
have texted while driving within the past 30 days?

1.  What is the proportion of people who have texted while driving every day in 
the past 30 days and never wear helmets?

Remember that you can use `filter` to limit the dataset to just non-helmet
wearers. Here, we will name the dataset `no_helmet`.

```{r no helmet}
no_helmet <- yrbss %>%
  filter(helmet_12m == "never")
```

Also, it may be easier to calculate the proportion if you create a new variable
that specifies whether the individual has texted every day while driving over the 
past 30 days or not. We will call this variable `text_ind`.

```{r indicator-texting}
no_helmet <- no_helmet %>%
  mutate(text_ind = ifelse(text_while_driving_30d == "30", "yes", "no"))
```

## Inference on proportions

When summarizing the YRBSS, the Centers for Disease Control and Prevention seeks 
insight into the population *parameters*. To do this, you can answer the question,
"What proportion of people in your sample reported that they have texted while 
driving each day for the past 30 days?" with a statistic; while the question 
"What proportion of people on earth have texted while driving each day for the 
past 30 days?" is answered with an estimate of the parameter.

The inferential tools for estimating population proportion are analogous to 
those used for means in the last chapter: the confidence interval and the 
hypothesis test.

1.  Write out the conditions for inference to construct a 95% confidence
    interval for the proportion of non-helmet wearers that have texted while
    driving for the past 30 days. Are you confident all conditions are met?

If the conditions for inference are reasonable, we can either calculate
the standard error and construct the interval by hand, or allow the `infer`
package to do it for us.

```{r nohelmet-text-ci, tidy = FALSE}
no_helmet %>%
  specify(formula = text_ind ~ NULL, success = "yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = .95)
```

Note that since the goal is to construct an interval estimate for a 
proportion, it's necessary to both include the `success` argument within `specify`,
which accounts for the proportion of non-helmet wearers than have consistently texted
while driving the past 30 days, in this example, and that `stat` within `calculate`
is here "prop", signaling that you are trying to do some sort of inference on a 
proportion.


1.  What is the margin of error for the estimate of the proportion of non-helmet 
wearers that have texted while driving each day for the past 30 days based on 
this survey?

1.  Using the `infer` package, calculate confidence intervals for the 
    proportion of atheists in 2012 in two other categorical variables, and 
    report the associated margins of error. Be sure to note whether the 
    conditions for inference are met, and interpet the interval in context of the
    data. It may be helpful to create new data sets for each of the two countries
    first, and then use these data sets to construct the confidence intervals.

## How does the proportion affect the margin of error?

Imagine you've set out to survey 1000 people on two questions: are you at least
6-feet tall? and are you left-handed? Since both of these sample proportions were 
calculated from the same sample size, they should have the same margin of 
error, right? Wrong! While the margin of error does change with sample size, 
it is also affected by the proportion.

Think back to the formula for the standard error: $SE = \sqrt{p(1-p)/n}$. This 
is then used in the formula for the margin of error for a 95% confidence 
interval: 
$$
ME = 1.96\times SE = 1.96\times\sqrt{p(1-p)/n} \,.
$$
Since the population proportion $p$ is in this $ME$ formula, it should make sense
that the margin of error is in some way dependent on the population proportion. 
We can visualize this relationship by creating a plot of $ME$ vs. $p$.

Since sample size is irrelevant to this discussion, let's just set it to
some value ($n = 1000$) and use this value in the following calculations:

```{r n-for-me-plot}
n <- 1000
```

The first step is to make a variable `p` that is a sequence from 0 to 1 with 
each number incremented by 0.01. You can then create a variable of the margin of 
error (`me`) associated with each of these values of `p` using the familiar 
approximate formula ($ME = 2 \times SE$).

```{r p-me}
p <- seq(from = 0, to = 1, by = 0.01)
me <- 2 * sqrt(p * (1 - p)/n)
```

Lastly, you can plot the two variables against each other to reveal their 
relationship. To do so, we need to first put these variables in a data frame that
you can call in the `ggplot` function.

```{r me-plot}
dd <- data.frame(p = p, me = me)
ggplot(data = dd, aes(x = p, y = me)) + 
  geom_line() +
  labs(x = "Population Proportion", y = "Margin of Error")
```

1.  Describe the relationship between `p` and `me`. Include the margin of
    error vs. population proportion plot you constructed in your answer. For
    a given sample size, for which value of `p` is margin of error maximized?

## Success-failure condition

We have emphasized that you must always check conditions before making 
inference. For inference on proportions, the sample proportion can be assumed 
to be nearly normal if it is based upon a random sample of independent 
observations and if both $np \geq 10$ and $n(1 - p) \geq 10$. This rule of 
thumb is easy enough to follow, but it makes you wonder: what's so special 
about the number 10?

The short answer is: nothing. You could argue that you would be fine with 9 or 
that you really should be using 11. What is the "best" value for such a rule of 
thumb is, at least to some degree, arbitrary. However, when $np$ and $n(1-p)$ 
reaches 10 the sampling distribution is sufficiently normal to use confidence 
intervals and hypothesis tests that are based on that approximation.

You can investigate the interplay between $n$ and $p$ and the shape of the 
sampling distribution by using simulations. Play around with the following
app to investigate how the shape, center, and spread of the distribution of
$\hat{p}$ changes as $n$ and $p$ changes.

```{r sf-app, echo=FALSE, eval=TRUE}
inputPanel(
  numericInput("n", label = "Sample size:", value = 300),
  
  sliderInput("p", label = "Population proportion:",
              min = 0, max = 1, value = 0.1, step = 0.01),
  
  numericInput("x_min", label = "Min for x-axis:", value = 0, min = 0, max = 1),
  numericInput("x_max", label = "Max for x-axis:", value = 1, min = 0, max = 1)
)

renderPlot({
  pp <- data.frame(p_hat = rep(0, 5000))
  for(i in 1:5000){
    samp <- sample(c(TRUE, FALSE), input$n, replace = TRUE, 
                   prob = c(input$p, 1 - input$p))
    pp$p_hat[i] <- sum(samp == TRUE) / input$n
  }
  bw <- diff(range(pp$p_hat)) / 30
  ggplot(data = pp, aes(x = p_hat)) +
    geom_histogram(binwidth = bw) +
    xlim(input$x_min, input$x_max) +
    ggtitle(paste0("Distribution of p_hats, drawn from p = ", input$p, ", n = ", input$n))
})
```

1.  Describe the sampling distribution of sample proportions at $n = 300$ and 
    $p = 0.1$. Be sure to note the center, spread, and shape.

1.  Keep $n$ constant and change $p$. How does the shape, center, and spread 
    of the sampling distribution vary as $p$ changes. You might want to adjust
    min and max for the $x$-axis for a better view of the distribution.
    
1.  Now also change $n$. How does $n$ appear to affect the distribution of $\hat{p}$?

* * *

## More Practice

For some of the exercises below, you will conduct inference comparing two 
proportions. In such cases, you have a response variable that is categorical, and
an explanatory variable that is also categorical, and you are comparing the 
proportions of success of the response variable across the levels of the 
explanatory variable. This means that when using `infer`, you need to include
both variables within `specify`.

1.  Is there convincing evidence that those who sleep 10+ hours per day are more
    likely to strength train every day of the week? As always, write out the 
    hypotheses for any tests you conduct and outline the status of the conditions
    for inference. If you find a significant difference, also quantify this 
    difference with a confidence interval. 

1.  Let's say there has been no difference in likeliness to strength train every
    day of the week for those who sleep 10+ hours. What is the probablity that
    you could detect a change (at a significance level of 0.05) simply by chance?
    *Hint:* Review the definition of the Type 1 error.

1.   Suppose you're hired by the local government to estimate the proportion of 
    residents that attend a religious service on a weekly basis. According to 
    the guidelines, the estimate must have a margin of error no greater than 
    1% with 95% confidence. You have no idea what to expect for $p$. How many 
    people would you have to sample to ensure that you are within the 
    guidelines?\
    *Hint:* Refer to your plot of the relationship between $p$ and margin of 
    error. This question does not require using a dataset.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
</div>
